### 2023-07-14 python_excel_ite.tex
working on last submodule

Python has continued to make inroads to analytical territory that was formerly dominated by spreadsheets. The widely used Python data analysis library, pandas, was created by Wes McKinney while working as a quantitative financial analyst \cite{PythonDataAnalysis}. Python is widely used for data wrangling, visualization and predictive modeling using various flavors of regression, tree based models and neural networks. The blog \href{https://pbpython.com/}{Practical Business Python} written primarily by Chris Moffitt has contributed to making Python more accessible to business analytics practitioners through high quality posts showing practical use of Python for business analysis problems. But, it got me to thinking. What about those things for which Excel is particularly well suited such as building formula based models and doing sensitivity analysis on these models? What would those look like in Python? The importance of modeling, articulated so well in the pages of this journal \cite{powellTeachingModelingManagement2001}, has not abated. What role does Python have to play in supporting the kind of modeling taught for years in our business schools?

@@

Since order quantity is the key decision variable, we might want to see how profit changes for different order quantities. In Excel, the Data Table tool provides an easy way to do this and is a staple of spreadsheet based management science textbooks. In a Data Table, a range of order quantities can be used as row or column input and one or more output values can be computed. In Python, the vectorized nature of the NumPy library provides a simple way to accomplish the same thing. A range of order quantities is created as a NumPy array that can be used directly to compute vectors of all of the intermediate and final output variables. Much like Data Tables, vectorized computations avoid having to explicitly iterate, or loop, through a collection of input values.

@@

In the Excel modeling world, Monte-Carlo simulation can be done without add-ins, but packages like @Risk \cite{} can make the process much easier. In particular, random variable generation for a wide range of probability distributions is facilitated by add-ins like @Risk. In Python, the NumPy package provide random variate generation for a large number of distributions and similarly, SciPy provides functions for computing distribution related quantities for many probability distributions \cite{}. Much like we did when creating the \code{data\_table} function, we start with adding uncertainty to a single input variable in the bookstore model and rely on Python's vectorized computing capabilities to generate the profit associated with each realization of the random input. Then we can do standard statistical analysis of the simulation output using Python packages such as pandas and SciPy. In spreadsheet based simulation textbooks, the notion of the ``Flaw of Averages'' popularized by Sam Savage \cite{} is often illustrated. We do the same with our Python based model, showing how replacing random demand by its mean can result in a wildly optimistic estimate of the mean profit found through analyzing the simulation output. Much like doing simulation in Excel without add-ins, using Python makes transparent the quite simple and brute force nature of Monte-Carlo simulation.

@@

This is followed by a well commented first attempt at a \code{simulate} function. By this point, students are well equipped to understand the code as it builds on basic Python concepts and more advanced concepts that we have already covered in this EwP module. The rest of the notebook is spent trying out the \code{simulate} function and ends with creating a grouped boxplot that many of them have likely created before with @Risk along with a faceted plot of histograms of profit created with the popular Seaborn package.

@@

\section{Submodule 2: The packaging notebooks}

@@

The first notebook in this submodule starts by explaining what Python packages are and their role in sharing code. We then revisit a concept from earlier in the course - creating a good project folder structure. For this we use what is known as a \code{cookiecutter} that can automatically generate a project folder structure and key files from the answers to a few prompts \cite{cookiecutter}. There are a few different folder layouts used in Python projects and there is a bit of discussion on the folder structure we will use and resources given for exploring this issue further. Students are reminded of the importance of version control and we initialize a git repository for our budding package. The basics of version control were already covered in the first week of the course. My experience is that most students have little exposure to software project management concepts such as project folder structures and version control. This can lead to a jumble of files with incoherent names used as a proxy for version control -- \code{whatif\_v1.py}, \code{whatif\_v2.py}, \code{whatif\_final.py}, \code{whatif\_finalfinal.py}, and so on.

@@

This notebook ends by pointing the students to another notebook in which our newly deployed whatif package is used for a completely different model. The example used is the New Car Simulation that has been a staple of Winston and Albright's textbooks for many years \cite{bibid}. This new model differs structurally from the Bookstore Model, but is representative of many financial models, in that cash flows over multiple years need to be computed and summarized. It provides a good test for our new whatif package.

@@

This deployment focused submodule ends with a short notebook discussing the different types of documentation needed in a typical project. This includes code comments, docstrings, readme files and generating documentation by writing restrutured text files and using Sphinx. This is one of those tasks that few people like to do but can be very important for the long term success of a project \cite{bibid}. Even if you are just creating a tool for yourself, you will forget the details over time and documentation can be very beneficial in refreshing your memory.

@@

A single Jupyter notebook is used to give a taste of using Python to automate the process of working with Excel files by tackling each of the four examples above. Each of these examples are based on a real problems I encountered either in research or industrial projects. Several web based resources are shared including the Practical Business Python blog \cite{bibid} written by Chris Moffitt, a data analytics professional. This blog, which focuses on helping Excel centric analytics professionals transition some of their work to Python, includes numerous posts on different aspects of Python and Excel integration.

@@

While this first example is not necessarily Excel related, using Excel to consolidate csv files through a series of manual Copy-Paste operations is something that many Excel users will admit to having done. After getting familiar with pathlib, a library for working with file systems which has recently been added to Python's standard library, students are led through the development of a short procedure that creates a list of the csv files to concatenate and then uses the pandas library to read and combine the files in a \code{DataFrame} and then writes out the consolidated data as a csv file.

@@

Two different approaches are shown to consolidate data from multiple sheets into a single sheet. One approach uses the pandas library while the other relies on a library specifically created for working with Excel files from Python, openpyxl \cite{bibid}. One advantage to using Python rather than Visual Basic for Applications for this task is that Python is well suited for working with data in many different formats including csv, json, xml and a myriad of database formats. The Excel file may be just one data source out of a collection of data sources that are used to create some sort of combined database. The generality of Python is a strength in such scenarios.

>>>>

Python has continued to make inroads to analytical territory that was formerly dominated by spreadsheets. The widely used Python data analysis library, pandas, was created by Wes McKinney while working as a quantitative financial analyst \cite{PythonDataAnalysis}. Python is widely used for data wrangling, visualization and predictive modeling using various flavors of regression, tree based models and neural networks. You can get a very good sense of this by visiting the very popular \href{https://www.reddit.com/r/datascience/}{r/datascience subreddit on Reddit} -- a well known online forum. But, it got me to thinking. What about those things for which Excel is particularly well suited such as building formula based models and doing sensitivity analysis on these models? What would those look like in Python? The importance of modeling, articulated so well in the pages of this journal \cite{powellTeachingModelingManagement2001}, has not abated. What role does Python have to play in supporting the kind of modeling taught for years in our business schools?

@@

Since order quantity is the key decision variable, we might want to see how profit changes for different order quantities. In Excel, the Data Table tool provides an easy way to do this and is a staple of spreadsheet based management science textbooks. In a Data Table, a range of order quantities can be used as row or column input and one or more output values can be computed. In Python, the vectorized nature of the \href{https://numpy.org/}{NumPy} library \cite{NumPy}provides a simple way to accomplish the same thing. A range of order quantities is created as a NumPy array that can be used directly to compute vectors of all of the intermediate and final output variables. Much like Data Tables, vectorized computations avoid having to explicitly iterate, or loop, through a collection of input values.

@@

In the Excel modeling world, Monte-Carlo simulation can be done without add-ins, but packages like @Risk \cite{RISKRiskAnalysis} can make the process much easier. In particular, random variable generation for a wide range of probability distributions is facilitated by add-ins like @Risk. In Python, the NumPy package provide random variate generation for a large number of distributions and similarly, SciPy provides functions for computing distribution related quantities for many probability distributions. Much like we did when creating the \code{data\_table} function, we start with adding uncertainty to a single input variable in the bookstore model and rely on NumPy's vectorized computing capabilities to generate the profit associated with each realization of the random input. Then we can do standard statistical analysis of the simulation output using Python packages such as pandas and SciPy. In spreadsheet based simulation textbooks, the notion of the ``Flaw of Averages'' popularized by Sam Savage \cite{savageFlawAveragesWhy2012} is often illustrated. We do the same with our Python based model, showing how replacing random demand by its mean can result in a wildly optimistic estimate of the mean profit found through analyzing the simulation output. Much like doing simulation in Excel without add-ins, using Python makes transparent the quite simple and brute force nature of Monte-Carlo simulation.

@@

This is followed by a well commented first attempt at a \code{simulate} function. By this point, students are well equipped to understand the code as it builds on basic Python concepts and more advanced concepts that we have already covered in the EwP module. The rest of the notebook is spent trying out the \code{simulate} function and ends with creating a grouped boxplot that many of them have likely created before with @Risk along with a faceted plot of histograms of profit created with the popular \href{https://seaborn.pydata.org/}{Seaborn} package.

@@

\section{Submodule 2: The packaging and documentation notebooks}

%TODO - Need transition opening from modeling to packaging for reusability

Don't want to be copying and pasting code for the next model. We have spent some effort in creating something that is reusable.
VBA for Modelers got into the deployment process in the Excel world.

Classroom based spreadsheet modeling problems often end here. The model is built and used to analyze a given problem.

@@

The first notebook in this submodule starts by explaining what Python packages are and their role in sharing code. We then revisit a concept from earlier in the course -- creating a good project folder structure. For this we use what is known as a \href{https://cookiecutter.readthedocs.io/en/stable/}{cookiecutter} that can automatically generate a project folder structure and key files from the answers to a few prompts \cite{CookiecutterBetterProject}. There are a few different folder layouts used in Python projects and there is a bit of discussion on the folder structure we will use and resources given for exploring this issue further. Students are reminded of the importance of version control and we initialize a git repository for our budding package. The basics of version control were already covered in the first week of the course. My experience is that most students have little exposure to software project management concepts such as project folder structures and version control. This can lead to a jumble of files with incoherent names used as a proxy for version control -- \code{whatif\_v1.py}, \code{whatif\_v2.py}, \code{whatif\_final.py}, \code{whatif\_finalfinal.py}, and so on.

@@

This notebook ends by pointing the students to another notebook in which our newly deployed whatif package is used for a completely different model -- the New Car Simulation that has been a staple of Winston and Albright's textbooks for many years \cite{winstonPracticalManagementScience2018,albrightBusinessAnalyticsData2016}. This new model differs structurally from the Bookstore Model, but is representative of many financial models in that cash flows over multiple years need to be computed and summarized.

@@

This deployment focused submodule ends with a short notebook discussing the different types of documentation needed in a typical project. This includes code comments, docstrings, \code{readme} files and generating documentation by writing \href{https://www.sphinx-doc.org/en/master/usage/restructuredtext/basics.html}{reStructuredText} files (similar but more powerful than markdown) and using \href{https://www.sphinx-doc.org/en/master/index.html}{Sphinx}, the Python documentation generation tool. This is one of those tasks that few people like to do but can be very important for the long term success of a project. Even if you are just creating a tool for yourself, you will forget the details over time and documentation can be very beneficial in refreshing your memory.

@@

A single Jupyter notebook is used to give a taste of using Python to automate the process of working with Excel files by tackling each of the four examples above. Each of these examples are based on a real problems I encountered either in research or industrial projects. Several web based resources are shared including the \href{https://pbpython.com/}{Practical Business Python blog} written by Chris Moffitt, a data analytics professional \cite{moffittPracticalBusinessPython}. This blog, which focuses on helping Excel centric analytics professionals transition some of their work to Python, includes numerous posts on different aspects of Python and Excel integration.

@@

While this first example is not necessarily Excel related, using Excel to consolidate csv files through a series of manual Copy-Paste operations is something that many Excel users will admit to having done. After getting familiar with pathlib, a library for working with file systems which has recently been added to Python's standard library, students are led through the development of a short procedure that creates a list of the csv files to concatenate and then uses pandas to read and combine the files in a \code{DataFrame} and writes out the consolidated data as a csv file.

@@

Two different approaches are shown to consolidate data from multiple sheets into a single sheet. One approach uses pandas while the other relies on a library specifically created for working with Excel files from Python, openpyxl \cite{bibid}. One advantage to using Python rather than Visual Basic for Applications for this task is that Python is well suited for working with data in many different formats including csv, json, xml and a myriad of database formats. The Excel file may be just one data source out of a collection of data sources that are used to create some sort of combined database. The generality of Python is a strength in such scenarios.
